maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-10-20 docs/paper/paper.tex
recompile paper

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\citeproc{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta, Bansal,
    et al., 2021}) and can be configured to be similar to the outputs of
    Németh (\citeproc{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\citeproc{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \citeproc{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\citeproc{ref-mazegenerator-net}{Alance AB, 2019};
    \citeproc{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \citeproc{ref-mathematica-maze}{Guo et al., 2011};
    \citeproc{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\citeproc{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}) but with many more options, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\citeproc{ref-zanj}{M. Ivanitskiy,
  n.d.}). As far as we are aware, no existing packages do this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\citeproc{ref-ayaz2008maze}{Ayaz et al., 2008})
  appear to simply no longer be accessible.
\end{itemize}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\citeproc{ref-deepthinking}{Schwarzschild, Borgnia, Gupta, Huang, et
al., 2021}). To facilitate the use of our package in this context, we
replicate the format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster}.

@@

Our package is implemented in Python(\citeproc{ref-python}{Rossum,
1995}), and makes use of the extensive scientific computing ecosystem,
including NumPy (\citeproc{ref-numpy}{Harris et al., 2020}) for array
manipulation, plotting tools (\citeproc{ref-matplotlib}{Hunter, 2007};
\citeproc{ref-seaborn}{Waskom, 2021}), Jupyter notebooks
(\citeproc{ref-jupyter}{Kluyver et al., 2016}), and PySR
(\citeproc{ref-pysr}{Cranmer, 2023}) for symbolic regression.

@@

This package was originally built for the needs of the
\texttt{maze-transformer} project
(\citeproc{ref-maze-transformer-github}{Michael I. Ivanitskiy et al.,
2023b}), which aims to investigate spatial planning and world models in
autoregressive transformer models trained on mazes
(\citeproc{ref-ivanitskiy2023structuredworldreps}{Michael Igorevich
Ivanitskiy, Spies, et al., 2023};
\citeproc{ref-maze-dataset-arxiv-2023}{Michael Igorevich Ivanitskiy,
Shah, et al., 2023}; \citeproc{ref-spies2024causalworldmodels}{Spies et
al., 2024}). It was extended for work on understanding the mechanisms by
which recurrent convolutional and implicit networks
(\citeproc{ref-fung2022jfb}{Fung et al., 2022}) solve mazes given a
rasterized view (\citeproc{ref-knutson2024logicalextrapolation}{Knutson
et al., 2024}), which required matching the pixel-padded and endpoint
constrained output format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}). Ongoing work using
\texttt{maze-dataset} aims to investigate the effects of varying the
tokenization format on the performance of pretrained LLMs on spatial
reasoning.

@@

\begin{itemize}
\item
  By (\citeproc{ref-nolte2024multistep}{Nolte et al., 2024}) to compare
  the effectiveness of transformers trained with the MLM-\(\mathcal{U}\)
  (\citeproc{ref-MLMU-kitouni2024factorization}{Kitouni et al., 2024})
  multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\citeproc{ref-wang2024imperative}{Wang et al., 2024}) and
  (\citeproc{ref-chen2024iaimperative}{Chen et al., 2024}) to study
  imperative learning.
\item
  By (\citeproc{ref-zhang2025tscend}{Zhang et al., 2025a}) to introduce
  a novel framework for reasoning diffusion models.
\item
  By (\citeproc{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to improve
  spatial reasoning in LLMs with GRPO.
\item
  By (\citeproc{ref-cai2025morse}{Cai et al., 2025}) to create a
  multimodal reasoning benchmark, via mazes in videos.
\item
  By (\citeproc{ref-xu2025visual}{Xu et al., 2025}) to study visual
  planning in LLMs.
\item
  By (\citeproc{ref-lee2025adaptive}{Lee et al., 2025}) to evaluate
  adaptive inference-time scaling with diffusion models on maze
  navigation tasks.
\item
  By (\citeproc{ref-zhang2025vfscale}{Zhang et al., 2025b}) to test
  verifier-free diffusion models.
\end{itemize}

@@

\bibitem[\citeproctext]{ref-pysr}
Cranmer, M. (2023). Interpretable machine learning for science with PySR
and SymbolicRegression. jl. \emph{arXiv Preprint arXiv:2305.01582}.
\url{https://doi.org/10.48550/arXiv.2305.01582}

>>>>

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\citeproc{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta, Bansal,
    et al., 2021}) and can be configured to be similar to the outputs of
    Németh (\citeproc{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to Singla
    (\citeproc{ref-eval-gpt-visual}{2023}) and Oppenheim
    (\citeproc{ref-gh-oppenheimj2018maze}{2018})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature-rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of many packages (\citeproc{ref-mazegenerator-net}{Alance
    AB, 2019}; \citeproc{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \citeproc{ref-mathematica-maze}{Guo et al., 2011};
    \citeproc{ref-mdl-suite}{Nag, 2020}).
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of Liu \& Wu (\citeproc{ref-eval-LLM-graphs}{2023})
  but with many more options, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\citeproc{ref-zanj}{M. Ivanitskiy,
  n.d.}). As far as we are aware, no existing packages do this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\citeproc{ref-ayaz2008maze}{Ayaz et al., 2008})
  appear to simply no longer be accessible.
\end{itemize}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\citeproc{ref-deepthinking}{Schwarzschild, Borgnia, Gupta, Huang, et
al., 2021}). To facilitate the use of our package in this context, we
replicate the format of Schwarzschild, Borgnia, Gupta, Bansal, et al.
(\citeproc{ref-easy_to_hard}{2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster}.

@@

Our package is implemented in Python (\citeproc{ref-python}{Rossum,
1995}), and makes use of the extensive scientific computing ecosystem,
including NumPy (\citeproc{ref-numpy}{Harris et al., 2020}) for array
manipulation, plotting tools (\citeproc{ref-matplotlib}{Hunter, 2007};
\citeproc{ref-seaborn}{Waskom, 2021}), Jupyter notebooks
(\citeproc{ref-jupyter}{Kluyver et al., 2016}), and PySR
(\citeproc{ref-pysr}{Cranmer, 2023}) for symbolic regression.

@@

This package was originally built for the needs of the
\texttt{maze-transformer} project
(\citeproc{ref-maze-transformer-github}{Michael I. Ivanitskiy et al.,
2023b}), which aims to investigate spatial planning and world models in
autoregressive transformer models trained on mazes
(\citeproc{ref-ivanitskiy2023structuredworldreps}{Michael Igorevich
Ivanitskiy, Spies, et al., 2023};
\citeproc{ref-maze-dataset-arxiv-2023}{Michael Igorevich Ivanitskiy,
Shah, et al., 2023}; \citeproc{ref-spies2024causalworldmodels}{Spies et
al., 2024}). It was extended for work on understanding the mechanisms by
which recurrent convolutional and implicit networks
(\citeproc{ref-fung2022jfb}{Fung et al., 2022}) solve mazes given a
rasterized view (\citeproc{ref-knutson2024logicalextrapolation}{Knutson
et al., 2024}), which required matching the pixel-padded and endpoint
constrained output format of Schwarzschild, Borgnia, Gupta, Bansal, et
al. (\citeproc{ref-easy_to_hard}{2021}). Ongoing work using
\texttt{maze-dataset} aims to investigate the effects of varying the
tokenization format on the performance of pretrained LLMs on spatial
reasoning.

@@

\begin{itemize}
\item
  By Nolte et al. (\citeproc{ref-nolte2024multistep}{2024}) to compare
  the effectiveness of transformers trained with the MLM-\(\mathcal{U}\)
  (\citeproc{ref-MLMU-kitouni2024factorization}{Kitouni et al., 2024})
  multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By Wang et al. (\citeproc{ref-wang2024imperative}{2024}) and Chen et
  al. (\citeproc{ref-chen2024iaimperative}{2024}) to study imperative
  learning.
\item
  By Zhang et al. (\citeproc{ref-zhang2025tscend}{2025a}) to introduce a
  novel framework for reasoning diffusion models.
\item
  By Dao \& Vu (\citeproc{ref-dao2025alphamaze}{2025}) to improve
  spatial reasoning in LLMs with GRPO.
\item
  By Cai et al. (\citeproc{ref-cai2025morse}{2025}) to create a
  multimodal reasoning benchmark, via mazes in videos.
\item
  By Xu et al. (\citeproc{ref-xu2025visual}{2025}) to study visual
  planning in LLMs.
\item
  By Lee et al. (\citeproc{ref-lee2025adaptive}{2025}) to evaluate
  adaptive inference-time scaling with diffusion models on maze
  navigation tasks.
\item
  By Zhang et al. (\citeproc{ref-zhang2025vfscale}{2025b}) to test
  verifier-free diffusion models.
\end{itemize}

@@

\bibitem[\citeproctext]{ref-pysr}
Cranmer, M. (2023). Interpretable machine learning for science with PySR
and SymbolicRegression.jl. \emph{arXiv Preprint arXiv:2305.01582}.
\url{https://doi.org/10.48550/arXiv.2305.01582}
